---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rsvddpd

<!-- badges: start -->
[![Build Status](https://www.travis-ci.com/subroy13/rsvddpd.svg?branch=master)](https://www.travis-ci.com/subroy13/rsvddpd)
<!-- badges: end -->

The `R` package `rsvddpd` is an acronym for Robust Singular Value Decomposition using Density Power Divergence. As the name suggests, the package mainly concerns with a special function for performing SVD in a robust way in presence of outliers. The details of the algorithm can be found in the paper [https://arxiv.org/abs/2109.10680](https://arxiv.org/abs/2109.10680).

There are 3 primary functions in the package.

* rSVDdpd - Performs the robust SVD of a numeric matrix $X$.
* simSVD - Simulates different scenarios to compare performances of SVD algorithms under outlier contamination.
* cv.alpha - It computes the optimal robustness parameter $\alpha$ for the `rSVDdpd` algorithm based on the data matrix $X$.


## Installation

You can install the development version from [GitHub](https://github.com/subroy13/rsvddpd) with:

``` r
# install.packages("devtools")
devtools::install_github("subroy13/rsvddpd")
```

Use the following to install the development version with manuals and vignettes, which provides useful information about the structure of the function.

``` r
devtools::install_github("subroy13/rsvddpd", build_opts = c("--no-resave-data"), build_manual = TRUE, build_vignettes = TRUE)
```


## Examples

This is a basic example usages which shows the need for the package.

```{r}
library(rsvddpd)

X <- matrix(1:20, nrow = 4, ncol = 5)
svd(X)
```

As you can see, the first two singular values are $53.5$ and $2.36$, and the third and fourth singular values are very small positive reals.

Let us see what happens when you contaminate just one entry of the matrix by a large value say $100$.

```{r}
X[2, 3] <- 100
svd(X)
```

Note that, the first singular value changes drastically, being $107$, while second and third singular values $36.1$ and $2.2$ respectively. However, such error is very common in practice, and can pose serious problem in many statistical estimation techniques. `rSVDdpd` solves the problem as shown in the following code.

```{r}
rSVDdpd(X, alpha = 0.3)
```

# Author

- Subhrajyoty Roy, Indian Statistical Institute, Kolkata (Package author and Maintainer)
- Ayanendranath Basu, Indian Statistical Institute, Kolkata
- Abhik Ghosh, Indian Statistical Institute, Kolkata


# Getting help

If you encounter a clear bug, please file an issue with a minimal reproducible example on [GitHub](https://github.com/subroy13/rsvddpd/issues).

-----

This package is distributed under [MIT](https://github.com/subroy13/rsvddpd/blob/master/LICENSE.md) license.



